Testing Gravity Using Type Ia Supernovae Discovered by Next Generation Wide Field Imaging Surveys

### 2019-01-19 doc/article.tex
plowing on

In the next decade, SNe~Ia will be used as peculiar-velocity probes to measure the the influence of gravity on structure formation within the Universe.
Peculiar velocities are sensitive to the growth of structure through the combination $fD$, where $D$ is  the linear growth factor and
$f \equiv \frac{d\ln{D}}{d\ln{a}}$ is the linear growth rate.  The growth of structure depends on gravity, indeed 
\citet{2007APh....28..481L} find that General Relativity, $f(R)$,  and DGP gravities 
give
$f \approx \Omega_M^\gamma$ with $\gamma=0.55, 0.42, 0.68$ respectively.  Adopting this parameterization for gravity, peculiar velocity
surveys are  sensitive to $fD=\Omega_M^{\gamma} \exp{\left(\int_a^1 \Omega_M^{\gamma} d\ln{a} \right)}$ \citep[see Figure~2 of][to see
$fD$ plotted as a function of redshift for $\Lambda$CDM]{1475-7516-2013-04-031}.

Peculiar velocity surveys have already been  used to measure $fD$, though not at a precision to distinguish gravity models.
 \citet{2017MNRAS.471..839A} use 6dFGS peculiar velocities using the Fundamental Plane method and get distances to
get a 15\% uncertainty in $fD$ at $z\approx 0$ using density and velocity cross-correlations.
The upcoming 
TAIPAN survey will obtain Fundamental Plane galaxies with densities of $n_g \sim 10^{-3}h^3$\,Mpc$^{-3}$,
and the WALLABY+WNSHS surveys will obtain Tully-Fisher galaxies at densities $n_g \sim 2\times 10^{-2} - 10^{-4} h^3$\,Mpc$^{-3}$ from
$z=0-0.1$.
Covering 75\% of the sky to a maximum depth of $z=0.1$, these surveys combined are projected to have 3\% $fD$ uncertainties \citep{2017MNRAS.464.2517H}.
Existing SN~Ia samples
have been used to test and ultimately find spatial correlations in peculiar velocities that may be attributed to the growth of structure
\citep{2008MNRAS.389L..47A,2014MNRAS.444.3926J,2015JCAP...12..033H, 2017JCAP...05..015H}.
For reference, t $z>0.3$ redshifts DESI projects 10\% precision of $fD$ using RSD alone.
%Measurement of the velocity field using LSST-discovered
%SNe~Ia has been quantified by \citet{2011PhRvD..83d3004B,2017JCAP...01..060O}

Two advances in the upcoming decade will make SNe~Ia  important probes of $f\sigma_8$.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter SED model yields $\sigma_M \gtrsim 0.12$~mag absolute magnitude
dispersion.  However, SNe transmit more information than just the light-curve shape and single color used in current SN models.
Recent studies indicate that with the right data, SNe absolute
magnitudes can be calibrated to $\sigma_M \lesssim 0.08$ mag \citep[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}.  Though not yet
established, it is anticipated that such a reduction in intrinsic dispersion comes with a reduction in the magnitude bias correlated with host-galaxy properties
that is observed using current calibrations.
One such SN is worth $\gtrsim 25$ galaxies with 0.4~mag absolute magnitude uncertainty.
Secondly,  in the upcoming decade cadenced wide-field imaging surveys such as ZTF and LSST
  will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands, over the course of 10-years, LSST will find $\sim150,000$ $z<0.2$,  $\sim520,000$ $z<0.3$ SNe~Ia for which good light curves can be measured. This is a sample size with more galaxies at deeper redshifts than projected by WALLABY and TAIPAN.

A 10-year LSST SN survey can produce interesting $fD$ measurements in redshift bins, as shown in Table~1 of 
\citet{2017ApJ...847..128H}.
Projected percent uncertainties in $f\sigma_8$ from 
\citet{2017ApJ...847..128H}. RSD is from clustering, and RSD~+~PV is from joint clustering and peculiar velocities.
The cumulative column shows the effective uncertainty from combining its and all other shallower redshift bins.

%      $0.00<z<0.05$   & 66.3 & 13.9 & 13.9\\
%     $0.05<z<0.10$            & 24.6     &  7.3 & 6.5\\
%     $0.10<z<0.15$      & 14.8  & 5.8 & 4.3\\
%     $0.15<z<0.20$      & 10.6  & 5.0 & 3.3\\
%      $0.20<z<0.25$     & 8.3  & 4.4 & 2.6\\
%     $0.25<z<0.30$  & 6.8  &  4.0 & 2.2\\

The current sample of SNe~Ia has a low number density compared to Fundamental Plane and Tully-Fisher galaxies.
Nevertheless, their low intrinsic-magnitude uncertainties 
can provide peculiar velocities (expressed equivalently as peculiar magnitudes)
of their host galaxies \citep{2006PhRvD..73l3526H,2011ApJ...741...67D}.

\section{Peculiar Velocity Surveys are Sensitive to the Growth of Structure $\lowercase{f}D$}
In a peculiar-velocity survey of SN~Ia host galaxies, each galaxy has an observed redshift  and a SN-Ia-based distance.
The distance maps onto a cosmological redshift given a model for the background expansion $H_0$-free.
The  (radial) peculiar velocity is obtained from the difference due to the Doppler effect between the observed and cosmological redshifts.
The statistical properties of the peculiar velocities are summarized through their spatial covariance or  their power spectrum $P_{vv}$.

@@

\section{Probing Gravity}
Our scientific interest is in probing gravity.
The combination $fD$ is sensitive to gravity. The growth rate $f$ describes how fast overdensities increase, which
depends on how fast gravity makes mass move.
General Relativity predicts
$f \approx \Omega_M^\gamma$ with $\gamma=0.55$, whereas other gravity models can be similarly described
with different values of
$\gamma=0.42, 0.68$ for $f(R)$,  and DGP gravities respectively 
\citep{2007APh....28..481L}.

As a probe of gravity, peculiar velocity surveys are sensitive to $fD=\Omega_M^{\gamma} a \exp{\left(\left(\int_0^a \Omega_M^{\gamma} -1\right)  d\ln(a)\right)}$.
Note that $\frac{d(\ln{fD})}{d\gamma} = \ln\Omega + \int \Omega^{\gamma} \ln{\Omega} d\ln{a} \approx -1.68, -0.75,-0.37$ at $z=0,0.5,1.0$ in $\Lambda$CDM,
meaning that surveys at low redshift better measure $\gamma$ given a fixed precent uncertainty in $fD$.
Non-GR models may  also predict a change in the scale-dependence of the growth, such observations provide additional leverage in probing gravity
but are not considered here.

The uncertainties in $\gamma$ for  RSD and peculiar velocity surveys are estimated using a
Fisher matrix analysis
similar to but with lower fidelity to that of \citet{2017ApJ...847..128H, 2017MNRAS.464.2517H}.
The analysis is performed in Fourier space, 
where the Fisher information matrix of parameters $\lambda \in \{\gamma,b\}$ from random Gaussian field with mean zero is given by
\begin{align}
F_{ij} & = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right]\\
& = \frac{\Omega}{8\pi^2} \int_0^{r_{max}}  \int_{k_{min}}^{k_{max}}  \int_{-1}^{1} r^2 k^2 \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right] d\mu\,dk\,dr
\label{fisher:eqn}
\end{align}
 and covariance
\begin{equation}
C(k,\mu)  =
  \begin{bmatrix}
   P_{gg}(k,\mu) + \frac{1}{n_g} &
   P_{vg}(k,\mu)  \\
   P_{vg}(k,\mu)  &
  P_{vv}(k,\mu) + \frac{\sigma^2}{n}
   \end{bmatrix}.
\label{cov:eq}
\end{equation}
The uncertainty in $\gamma$ is $\sqrt{\left(F^{-1}\right)_{\gamma,\gamma}}$.

%The growth of structure depends on the expansion history of the Universe, the nature and density of its contents, and gravity. 
%It is therefore a powerful probe of cosmology and dark energy.  Growth of structure can be measured from the light-emittting structures
%in the Universe and from the peculiar velocities of test masses therein.
%Peculiar velocities are the motions, on top of the cosmological expansion, caused by the gravitational attraction
%and repulsion of density inhomogeneities in the Universe.  The peculiar velocity of an object with a known absolute magnitude
%can be determined from its observer-frame magnitude and redshift. For a given background cosmology, the observed magnitude provides
%an estimate of the cosmological redshift, the peculiar velocity is then the difference between the cosmological and CMB-frame redshifts.

>>>>

In the next decade, SNe~Ia will be used as peculiar-velocity probes to measure the the influence of gravity on structure formation within the Universe.
The peculiar velocity power spectrum is sensitive to the growth of structure as $P_{vv}\propto (fD)^2$, where $D$ is  the linear growth factor and
$f \equiv \frac{d\ln{D}}{d\ln{a}}$ is the linear growth rate  \citep{2006PhRvD..73l3526H,2011ApJ...741...67D}. 
The $\Lambda$CDM prediction for the $z=0$ velocity power spectrum is shown in Figure~\ref{power:fig}. The growth of structure depends on gravity, indeed 
\citet{2007APh....28..481L} find that General Relativity, $f(R)$,  and DGP gravities 
give
$f \approx \Omega_M^\gamma$ with $\gamma=0.55, 0.42, 0.68$ respectively.  Adopting this parameterization for gravity, peculiar velocity
surveys are  sensitive to $fD=\Omega_M^{\gamma} \exp{\left(\int_a^1 \Omega_M^{\gamma} d\ln{a} \right)}$ \citep[see Figure~2 of][to see
$fD$ plotted as a function of redshift for $\Lambda$CDM]{1475-7516-2013-04-031}.

Peculiar velocity surveys have already been  used to measure $fD$, though not at a precision to distinguish gravity models.
 \citet{2017MNRAS.471..839A} use 6dFGS peculiar velocities using the Fundamental Plane method and get distances to
get a 15\% uncertainty in $fD$ at $z\approx 0$ using density and velocity cross-correlations.
The upcoming 
TAIPAN survey will obtain Fundamental Plane galaxies with densities of $n_g \sim 10^{-3}h^3$\,Mpc$^{-3}$,
and the WALLABY+WNSHS surveys will obtain Tully-Fisher galaxies at densities $n_g \sim 2\times 10^{-2} - 10^{-4} h^3$\,Mpc$^{-3}$ from
$z=0-0.1$.
Covering 75\% of the sky to a maximum depth of $z=0.1$, these surveys combined are projected to have 3\% $fD$ uncertainties \citep{2017MNRAS.464.2517H}.
Existing SN~Ia samples
have been used to test and ultimately find spatial correlations in peculiar velocities that may be attributed to the growth of structure
\citep{2008MNRAS.389L..47A,2014MNRAS.444.3926J,2015JCAP...12..033H, 2017JCAP...05..015H}.
For reference, DESI projects 10\% precision of $fD$ at $z>0.3$ redshifts  using RSD alone.
%Measurement of the velocity field using LSST-discovered
%SNe~Ia has been quantified by \citet{2011PhRvD..83d3004B,2017JCAP...01..060O}

Two advances in the upcoming decade will make SN~Ia peculiar velocities more powerful.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter SED model yields $\sigma_M \gtrsim 0.12$~mag absolute magnitude
dispersion.  However, SNe transmit more information than just the light-curve shape and single color used in current SN models.
Recent studies indicate that with the right data, SNe absolute
magnitudes can be calibrated to $\sigma_M \lesssim 0.08$ mag \citep[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}.  Though not yet
established, it is anticipated that such a reduction in intrinsic dispersion comes with a reduction in the magnitude bias correlated with host-galaxy properties
that is observed using current calibrations.
One such SN is worth $\gtrsim 25$ galaxies with 0.4~mag absolute magnitude uncertainty.
Secondly,  in the upcoming decade cadenced wide-field imaging surveys such as ZTF and LSST
  will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands, over the course of 10-years, LSST will find $\sim150,000$ $z<0.2$,  $\sim520,000$ $z<0.3$ SNe~Ia
with number density  $n \sim 5\times 10^{-4}h^3$\,Mpc$^{-3}$
 for which good light curves can be measured. This sample has comparable
 number density and more galaxies at deeper redshifts than projected by WALLABY and TAIPAN.

For example, over the course of a decade an LSST SN survey can produce interesting $fD$ measurements in several
redshift bins.   
\citet{2017ApJ...847..128H} find 4--14\% uncertainties in $fD$ in $0.05$ redshift bins from $z=0$ to 0.3, where at
$0<z<0.2$ most of the signal comes from peculiar velocities and at higher redshifts $RSD$ is more constraining.

Our scientific interest is in probing gravity, so while most of the literature cast peculiar velocity results in terms of 
$fD$, the interesting parameter is $\gamma$.  Survey design should be tailored toward minimizing the uncertainty
in 
$\sigma_\gamma$ rather than $\sigma_{\ln{fD}}$.  To illustrate the distinction,
 $\frac{d(\ln{fD})}{d\gamma} = \ln\Omega + \int \Omega^{\gamma} \ln{\Omega} d\ln{a}$, which for $\Lambda$CDM
 $\approx -1.68, -0.75,-0.37$ at $z=0,0.5,1.0$ respectively; given two surveys with the same percentage precision  in $fD$, the one at lower redshift
better constrains $\gamma$.  We thus thus relate surveys to $\sigma_\gamma$, where we assume a constant $\gamma$
over the redshift range of the survey.

%
%The combination $fD$ is sensitive to gravity. The growth rate $f$ describes how fast overdensities increase, which
%depends on how fast gravity makes mass move.
%General Relativity predicts
%$f \approx \Omega_M^\gamma$ with $\gamma=0.55$, whereas other gravity models can be similarly described
%with different values of
%$\gamma=0.42, 0.68$ for $f(R)$,  and DGP gravities respectively 
%\citep{2007APh....28..481L}.

The uncertainties in $\gamma$ are estimated using a
Fisher matrix analysis
similar  to that of \citet{2017ApJ...847..128H, 2017MNRAS.464.2517H}.
The analysis is performed in Fourier space, 
where the Fisher information matrix of parameters $\lambda \in \{\gamma,bD\}$ with
the  tracer density and velocity power spectra used as data is
\begin{align}
F_{ij} 
%& = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
%\frac{\partial C}{\partial \lambda_j} \right]\\
& = \frac{\Omega}{8\pi^2} \int_0^{r_{max}}  \int_{k_{min}}^{k_{max}}  \int_{-1}^{1} r^2 k^2 \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right] d\mu\,dk\,dr
\label{fisher:eqn}
\end{align}
where
\begin{equation}
C(k,\mu)  =
  \begin{bmatrix}
   P_{\delta \delta}(k,\mu) + \frac{1}{n} &
   P_{v\delta}(k,\mu)  \\
   P_{v\delta}(k,\mu)  &
  P_{vv}(k,\mu) + \frac{\sigma^2}{n}
   \end{bmatrix}.
\label{cov:eq}
\end{equation}
The parameter dependence comes in through $P_{vv}\propto ((fD)(\gamma)\mu)^2$, the SN~Ia power spectrum $P_{\delta \delta }\propto ((bD) + (fD)(\gamma)\mu^2)^2$, and the galaxy-velocity cross-correlation $P_{vg}
\propto  ((bD) + (fD)\mu^2)f\sigma_8$, where $b$ is the galaxy bias and $\mu$ is the cosine of the angle between the $k$-mode
and the line of sight.

The uncertainty in $\gamma$ is $\sqrt{\left(F^{-1}\right)_{\gamma,\gamma}}$.
Non-GR models may  also predict a change in the scale-dependence of the growth or deviations
from $f=\Omega_M^{\gamma}$, such observations provide additional leverage in probing gravity
but are not considered here.

%
%
%The current sample of SNe~Ia has a low number density compared to Fundamental Plane and Tully-Fisher galaxies.
%Nevertheless, their low intrinsic-magnitude uncertainties 
%can provide peculiar velocities (expressed equivalently as peculiar magnitudes)
%of their host galaxies \citep{2006PhRvD..73l3526H,2011ApJ...741...67D}. 
%
\section{Peculiar Velocity Surveys are Sensitive to the Growth of Structure $\lowercase{f}D$}
In a peculiar-velocity survey of SN~Ia host galaxies, each galaxy has an observed redshift  and a SN-Ia-based distance.
The distance maps onto a cosmological redshift given a model for the background expansion $H_0$-free.
The  (radial) peculiar velocity is obtained from the difference due to the Doppler effect between the observed and cosmological redshifts.
The statistical properties of the peculiar velocities are summarized through their spatial covariance or  their power spectrum $P_{vv}$.

@@

\section{Probing Gravity}
%The growth of structure depends on the expansion history of the Universe, the nature and density of its contents, and gravity. 
%It is therefore a powerful probe of cosmology and dark energy.  Growth of structure can be measured from the light-emittting structures
%in the Universe and from the peculiar velocities of test masses therein.
%Peculiar velocities are the motions, on top of the cosmological expansion, caused by the gravitational attraction
%and repulsion of density inhomogeneities in the Universe.  The peculiar velocity of an object with a known absolute magnitude
%can be determined from its observer-frame magnitude and redshift. For a given background cosmology, the observed magnitude provides
%an estimate of the cosmological redshift, the peculiar velocity is then the difference between the cosmological and CMB-frame redshifts.
